\pkg{digiRhythm}: an \proglang{R} package for rhythmicity assessment using the degree of functional coupling

### 2023-06-06 jstatsoft/article.Rnw
New version with Marianne/Hassan's and Marie discussions in seperate sections

\section{Discussion} \label{sec:summary}
The \pkg{digiRhythm} package can make a significant contribution to the assessment of rhythmicity, as it makes the analysis of rhythmicity more accessible to researchers. The tool is easy to use, especially because it is delivered as an \proglang{R} package. The theoretical details and intuition behind the DFC is explained in section \ref{sec:dfc}. The usage of the package is explained in the section \ref{sec:usage}. Further examples can be found in the library's vignettes.

%package enables computation of any activity data
The \pkg{digiRhythm} package enables the calculation of rhythmicity from any activity data. This indicator can be calculated from any type of regularly measured time series data, if the data are prepared in a suitable format. The suitability of the format can be assessed with the  \fct{is\_dgm\_friendly} function. To date, this is purely a computation of activity patterns, resulting in the DFC. This measure has been correlated with physiological functions in previous studies.

% Scope of DFC application
The computation of rhythmicity allows to gain further insight into an individual's harmonic patterns. Individuals in a good physiological state will present harmonic locomotory patterns, whereas this rhythmicity desynchronizes as a result of physiological events. Simply spoken, an organism will follow the same patterns of locomotion as long as they are in a good physiological and psychological state, whereas these patterns will deviate in the face of dysfunctions. In the past these rhythmic patterns have been discussed with regard to the psychological state of a humans, where psychological challenges, such as bipolar disorders or depression have been associated with the desynchronization of rhythmic patterns (\cite{sabet2021the}).
%Limitation of DFC computation

However, we have no evidence if these activity patterns are intrinsically (such as going to sleep) or extrinsically (such as waking up to an alarm) motivated. Such extrinsic factors could affect rhythmic activity patterns and blur physiological rhythmicity patterns. To date, the package provides the option to exclude extrinsically structured time spans from the analysis. However, in order to increase the understanding of rhythmicity  future studies should investigate how these extrinsic factors affect rhythmicity calculations.

The required frequency of data collection is of utmost importance in sensor technologies. Sensors are required to be small and further expected to run without any sensor based maintenance, such as data reading or charging of the device. This in turn means that data collection frequency should be low. In the current study a low data collection frequency is suggested (15 minutes) future studies should evaluate how low collection frequency can be set to create reliable rhythmicity calculations. Further it would be interesting to evaluate if activity data collected from different sources, such as  access of feeding stations or coffee maschines, compute similar rhythmicity outputs in the same organism. Seven days of continuous data collection are required to compute one DFC value. This implies that very long data collection intervals are required to detect rhythmic patterns. Reducing the number of days required to compute one DFC value would make this rhythmicity computation much more accessible in practice and research. However, the affect of reducing the intervals are not investigated, yet. Therefore, the impact of reducing the number of days required to compute the DFC should be investigated in future research.

In addition to the open questions about the data collection frequency and intervals, a comparison to other metrics will be needed in further studies. In the past, other metrics for calculating circadian rhythms have been published. For example, Interdaily Stability (IS), which is mainly used in human research \cite{witting1990alterations}. And the Fourier-Based Approximation with Thresholding (FBAT) was designed to calculate the circadian rhythms of dairy cows \cite{wagner2021detection}. Their database depends on the 24 h day, but none of them uses the 24 h within the calculation. The calculation of harmonic frequencies within the DFC calculation, which sets the activity signal in relation to the external 24h day is unique. Therefore, it would be interesting to investigate the correlation between the DFC and rhythmicity computed on data from various computation techniques. In addition to the mentioned metrics (IS and FBAT), \cite{moller2022diurnal} utilized a range of techniques, including set logic, circular cross-correlation, circular clustering, functional enrichment analyses, and least squares regression, on transcriptomic time series data from humans and baboons. Additionally, GPS data of free ranging animals \cite{plaza2022gps} or data of real time positioning systems in a barn \cite{wagner2021detection} could be used to estimate rhythmicity. Comparing the correlations between rhythmicity measures derived from different types of data and between various techniques could be a useful research area. In addition, further studies will be conducted, to validate the DFC calculation, by using simulated data.

The current package has been developed for farm animals, but the principles of rhythmicity apply for any organism, including humans. Therefore we highly recommend future research to evaluate the \pkg{digiRhythm} package in other species. Many humans nowadays continuously collect activity data with their smart-watches or their phones. Thus, it would be interesting to evaluate the application of the \pkg{digiRhythm} package in humans, offering an objective indicator of human welfare. A previous review highlights the potential of such rhythmicity computations in humans \citep{dijk2020novel}.

At the time of writing this paper, the current version (v1.2) of the library does not provide the capability to compute rhythmicity using techniques such as cosinor analysis (\cite{refinetti2016}). However, there is potential for future expansion of the library's functionality to incorporate additional techniques. Researchers are encouraged to actively contribute by submitting new data, codes, or methods to the GitHub repository. These contributions will be reviewed by the package maintainers and, if approved, integrated into the \pkg{digiRhythm} package.

\section{Discussion (New proposal Hassan)}

The \pkg{digiRhythm} package presents a novel, user-friendly, and accessible tool for researchers in the field of Rhythmicity, animal behavior, and beyond. The package offers a versatile approach to assessing rhythmicity based on digital activity records. By leveraging the power of \proglang{R}, it brings robust statistical capabilities and intuitive visualization features into the hands of researchers.

@@

>>>>

\section{Discussion} \label{sec:summary}
% Scope of DFC application
The \pkg{digiRhythm} package presents a novel, user-friendly, and accessible tool for researchers in the field of rhythmicity, animal behavior, and beyond. The package offers a versatile approach to assessing rhythmicity based on digital activity records. By leveraging the power of \proglang{R}, it brings robust statistical capabilities and intuitive visualization features into the hands of researchers.

@@

%package enables computation of any activity data
Rhythmicity can be calculated in the \pkg{digiRhythm} package from any type of regularly measured time series data, if the data are prepared in a suitable format. The suitability of the format can be assessed with the  \fct{is\_dgm\_friendly} function. To date, this is purely a computation of activity patterns, resulting in the DFC. This measure has been correlated with physiological functions in previous studies. The computation of rhythmicity allows to gain further insight into an individual's harmonic patterns. Individuals in a good physiological state will present harmonic locomotion patterns, whereas this rhythmicity desynchronizes as a result of physiological events. Simply spoken, an organism will follow the same patterns of locomotion as long as they are in a good physiological and psychological state, whereas these patterns will deviate in the face of dysfunctions. In the past, these rhythmic patterns have been discussed with regard to the psychological state of humans, where  bipolar disorders or depression have been associated with the desynchronization of rhythmic patterns (\cite{sabet2021the}).

%Limitation of DFC computation
However, there is no evidence to whether these activity patterns are intrinsically (such as going to sleep) or extrinsically (such as waking up to an alarm) motivated. Such extrinsic factors could affect rhythmic activity patterns and blur the organisms intrinsic rhythmicity. To date, the package provides the option to exclude extrinsically structured time spans from the analysis. However, in order to increase the understanding of rhythmicity  future studies should investigate how these extrinsic factors affect rhythmicity calculations.
Future research should also address data collection frequencies. Frequency of data collection is of great importance in sensor technology, because sensors are required to be small and expected to run without any sensor based maintenance, such as data reading or charging. This in turn implies that data collection frequencies should be set to a minimum. Although the current package used low data collection frequency in its example data (15 minutes), this time was given by the preset of the sensor data evaluated with this package and has no scientific basis. Future studies should therefore evaluate how low data collection frequencies can be set to compute reliable rhythmicity indicators.

Additionally, seven days of continuous data collection are currently required to compute one DFC value, implying that long data collection intervals are necessary to detect rhythmic patterns. Hereby, the basis for choosing a 7 day window is unclear. Reducing the number of days required to compute one DFC value would make this rhythmicity computation more accessible. Yet, the effect of reducing the intervals has not been investigated. Therefore, we encourage future studies to investigate the impact of reducing the number of days required to compute the DFC.

In addition to the open questions about the data collection frequency and intervals, a comparison to other metrics will be needed in further studies. In the past, other metrics for calculating circadian rhythms have been published. For example, Interdaily Stability (IS), which is mainly used in human research \cite{witting1990alterations}. And the Fourier-Based Approximation with Thresholding (FBAT) was designed to calculate the circadian rhythms of dairy cows \cite{wagner2021detection}. Their database depends on the 24 h day, but none of them use the 24 h within the calculation. The calculation of harmonic frequencies within the DFC calculation, which sets the activity signal in relation to the external 24h day is unique. Therefore, it would be interesting to investigate the correlation between the DFC and rhythmicity computed on data from various computation techniques. In addition to the mentioned metrics (IS and FBAT), \cite{moller2022diurnal} utilized a range of techniques, including set logic, circular cross-correlation, circular clustering, functional enrichment analyses, and least squares regression, on transcriptomic time series data from humans and baboons. Additionally, GPS data of free ranging animals \cite{plaza2022gps} or data of real time positioning systems in a barn \cite{wagner2021detection} could be used to estimate rhythmicity. Comparing the correlations between rhythmicity measures derived from different types of data and between various techniques could be a useful research area. Further, research should evaluate if activity data collected from different sources, such as tri-axial activity data, access of feeding stations or coffee machines, compute similar rhythmicity outputs in the same organism. In addition, further studies usinf simulated data should be conducted, to validate the DFC calculation.

The current package has been developed for farm animals, but the principles of rhythmicity apply for any organism, including humans. Therefore we highly recommend future research to evaluate the \pkg{digiRhythm} package in other species. Many humans nowadays continuously collect activity data with their smart-watches or their phones. A previous review highlighted the potential of such rhythmicity computations in humans \citep{dijk2020novel}. Thus, it would be interesting to implement the \pkg{digiRhythm} package in this domain. F

At the time of writing this paper, the current version (v1.2) of the library does not provide the capability to compute rhythmicity using techniques such as cosinor analysis (\cite{refinetti2016}). However, there is potential for future expansion of the library's functionality to incorporate additional techniques.
